Stability of variational quantum dynamics with Gaussian approximants

### 2013-09-08 paper.tex
writing

There appear to be two sorts of instabilities.  The first, static, sort occurs when a quantum state is expanded over a finite set of coherent states, and those states have irregular amplitudes.  This is similar to the kind that numerical analysts call ill-posed problems, and the methods they use to analyse and resolve them can be applied here.  However, the cause of ill-posedness in our problem differs from those usually studied, and the usual methods of resolution don't work as well as they usually do.  There is one advantage of the expansion problem: because it is iterated at each timestep, the dynamics of the amplitudes can be exploited to so that they posed better expansion problems as the simulation progresses.

@@

@@

@@

\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert

@@

\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert

@@

\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
\XeTeXpicfile rsv4.pdf width 0.5\hsize}
\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.  These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert

@@

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.  The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.  The tension is the Frobenius norm of the matrix $T$ defined in the text.
\endinsert

@@

\topinsert\XeTeXpicfile circle-1.pdf 
\endinsert

@@

\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.

The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.  This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.  Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.  Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.  When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.  This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
\endinsert

@@

\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert

@@

\topinsert\XeTeXpicfile cats.pdf width \hsize
\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert

@@

\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
Picard plot for cat expansion. \endinsert

@@

\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert

\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert

@@

\topinsert \XeTeXpicfile geometry-1.pdf
\it A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.  As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.  This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
\endinsert

\topinsert \XeTeXpicfile cancel.pdf width \hsize
\it The vacuum state was expanded over two coherent states, with amplitudes $±h$.  The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.  The norm of the expansion vector, and the residual of the expansion, are graphed above as black lines.  As the amplitudes converge, the residual decreases—the leftmost phase space plot is very nearly exact—while the norm of the expansion stays around 1.  This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.

@@

If the two sample amplitudes were part of an irregular expansion, and there happened to be no other sample amplitudes nearby, a small component along the direction of the second state would dominate the expansion coefficients at these points.  But, given those sample amplitudes, that would be the only way to represent such a component.
\endinsert

@@

Let the singular value decomposition of $|A(α)〉$ be $|A(α)〉=|U〉SV†$.  The standard ways to compute singular values use the matrix
$$M(x,y)=\pmatrix{0 & |A'(x,y)〉\cr 〈A'(x,y)| & 0}.$$
Here, $|A(x,y)〉$ maps weight vectors in $C^R$ to kets, while $〈A(x,y)|=VS〈U|$ maps kets back to column vectors of inner products.  The operation of the zeros is obvious; formally, they need to be a row of null kets and a column of null bras.  The matrix $M(x,y)$ operates on a vector comprising a ket above $R$ complex numbers.  Now, the vector
$$x={1\over \sqrt2}\pmatrix{|u_R〉 \cr v_R}$$
is a normalised eigenvector  of $M$, belonging to the eigenvalue $σ_R(x,y)$.  As is well known in first-order pertubation theory, the derivative of this eigenvalue with respect to $x$ and $y$ is given by
$$∂_rσ_R=x†(∂_rM)x
	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA〉v_R\right).$$
Substituting the partial derivatives gives
$$∂_rσ_R=\cases{\Re\left(〈u_R|(-x_r+a†)|α_r〉v_R^r\right) &  $1≤ r≤ R$, \cr
	\Re\left(〈u_R|(-y_r+ia†)|α_r〉v_R^r\right) & $R<r≤ 2R$.}$$

@@

\topinsert 
\vskip -40mm
\XeTeXpicfile psfa.pdf width \hsize
\vskip -90mm
\XeTeXpicfile psv.pdf width \hsize\endinsert

@@

\centerline{\XeTeXpicfile j1.pdf width \hsize}

\centerline{\XeTeXpicfile j2.pdf width \hsize}

@@

\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert

@@

\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert

@@

\topinsert \XeTeXpicfile spsns.pdf width \hsize
Amplitudes of components \endinsert

\topinsert \XeTeXpicfile traj.pdf width \hsize
Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert

\topinsert \XeTeXpicfile picard.pdf width \hsize
Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert

\topinsert \XeTeXpicfile svnum.pdf width \hsize
Expected particle number for left singular vectors.  The svs with small sws have large particle number.  When $H$ is quartic, these have large components in $H|ψ〉$.  In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$. \endinsert

@@

\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert

\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert

@@

\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}

\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}

>>>>

The instability appears to have two major causes.  The first problem occurs when quantum states are expanded over a finite set of coherent states.  There is a natural way to expand a state as an integral over the coherent states, with a weight function that varies smoothly with the coherent amplitude.  When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.  Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number—for instance, it might be a coherent state.  When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.  This causes the expansion weights to be large, to not resemble samples of any smooth function, and to change sharply as the state changes.  All these properties make it difficult to numerically integrate a rate of change that has been expanded by least squares.  The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.

The behaviour of the discrete expansion weights is similar to that observed in discrete ill-posed problems, and the methods used to analyse and resolve these can be applied to expansions over coherent states.  However, the cause of ill-posedness in our problem differs from those usually studied—the large number components of the state are real, while the high-frequency components that play the same role in ill-posed problems are spurious ones introduced by errors in measured data.  Therefore the standard methods of resolution don't work as well as usual.  The expansion problem is easier in one way: it is iterated at each timestep, so the dynamics of the coherent amplitudes can be modified so that the expansion problem becomes better posed as the simulation progresses.

@@

\beginsection{Variational dynamics}

Schrödinger's equation, the quantum law of motion, is
$$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
I will discuss many details of mappings from complex numbers to quantum states, so some notation used in this paper is traditionally used more in mathematics than in physics.  The symbol $|Ψ〉$ denotes a ket-valued function of a real number; $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.  The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.  Therefore it is written outside the ket.  The total derivative $D$ acts on the function, so it is written inside the ket.  Because $|Ψ〉$ is a function of a single variable, $|DΨ〉$ is simply another ket-valued function, and $|DΨ(t)〉$ is the ket that would traditionally be written ${∂\over ∂t}|Ψ〉\big|_{t=t}$.  This could alternatively be written as a partial derivative $|DΨ〉=|∂₀Ψ〉$.  If $|Ψ〉$ were a function of several variables, the total derivative function $|DΨ〉$ would evaluate to a row vector $|DΨ(t)〉=\pmatrix{|∂₀Ψ(t)〉&⋯&|∂_RΨ(t)〉}$.  One can regard the kets as column vectors and read this as a Jacobian matrix with infinitely many rows and $R$ columns.  The notation will be elaborated below as needed.

Schrödinger's equation could even be abbreviated to
$$i\hbar |DΨ〉=H|Ψ〉.$$
This is a very simple equation.  The challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand it over a grid grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes merely thousands of particles.  In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
$$|Ψ(t)〉=|ψ(z(t))〉,$$
or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\left(z₁(t),…,z_R(t)\right)$ is a vector of complex parameters.  This imposes two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.

Some methods of computing dynamics, such as those involving Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.  The above model does not apply to these.  There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which can compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store all the parameters at once.

The next sections will discuss the approximant $|ψ(z)〉$ analysed in this paper.  Before that, I'll outline the method that will be used to compute the dynamical $z(t)$.  The final part of the paper discusses the details—and there are many—of making this computation numerically stable.

If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
$$i\hbar|Dψ(z(t))〉Dz(t)=H|ψ(z(t))〉.$$
On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.  This is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to have components outside this subspace.  This is a consequence of the restricted form of the approximation.  We can, however, look for a vector $\dot z(z)$ that minimises the residual
$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\eqno{(4)}$$
and solve the ordinary differential equation
$${dz\over dt}=\dot z(z),\tag\ode$$
giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.  When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.

The Dirac-Frenkel solution is only optimised locally.  There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkey solution $|ψ(w(t))〉$ does.  This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.  When we discuss regularisation below, we will see that this is in fact the usual case.  However, global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.

The residual above can be defined using any norm one wishes.  The aim of computing quantum dynamics is usually to find some expectation values that could be compared to observed quantities, and one could consider using a norm that magnified the discrepencies in these observables, so that the solution with the least residual would correspond closely to the exact solution in the directions of interest.  On the other hand, we are computing quantum dynamics: these are unitary in the usual norm, so that a small error introduced at an early time is guaranteed to stay small when the state is propagated to later times.  No other norm guarantees this: they all introduce uncontrolled errors regarding how much early errors are magnified by later dynamics.

Equation~\ode\ will be integrated numerically.  For this to be done stably and efficiently, $\dot z$ must be a reasonably smooth function of $z$.  This is formally expressed in terms of the eigenvalues of the Jacobian $D\dot z$; different integration formulae place different bounds on the magnitudes and phases of these, but they all have limited regions of stability.  Moreover, as we will see below, a technically stable integration formula might still not produce a useful solution.

This complicates the solution of the problem in Formula~4.  It does not suffice to find a $\dot z$ with the smallest residual, for a given value of $z$.  In order for Equation~5 to be numerically solvable, the properties of $\dot z$ under perturbation of $z$ are equally important.  Under the usual Hilbert space norm, Formula~4 is a least squares problem; the pertubation properties of such problems are reasonably well understood.  We will examine ways to modify the problem, by introducing extra least-squares constraints, that might improve its perturbation properties.

@@

The approximant studied in this paper has the form
$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
	=|A〉d,$$
where $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.  In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.  The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion $|A〉$.  Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.  This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.

@@

%\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert

@@

%\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert

@@

%\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
%\XeTeXpicfile rsv4.pdf width 0.5\hsize}
%\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
%The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.  These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert

@@

%\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.  The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.  The tension is the Frobenius norm of the matrix $T$ defined in the text.
%\endinsert

@@

%\topinsert\XeTeXpicfile circle-1.pdf \endinsert

@@

%\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.

The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.  This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.  Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.  Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.  When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.  This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
%\endinsert

@@

%\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert

@@

%\topinsert\XeTeXpicfile cats.pdf width \hsize
%\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert

@@

%\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
%Picard plot for cat expansion. \endinsert

@@

%\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert

%\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert

@@

%\topinsert \XeTeXpicfile geometry-1.pdf
A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.  As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.  This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
%\endinsert

%\topinsert \XeTeXpicfile cancel.pdf width \hsize
The vacuum state was expanded over two coherent states, with amplitudes $±h$.  The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.  The norm of the expansion vector, and the residual of the expansion, are graphed above as black lines.  As the amplitudes converge, the residual decreases—the leftmost phase space plot is very nearly exact—while the norm of the expansion stays around 1.  This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.

@@

If the two sample amplitudes were part of an irregular expansion, and there happened to be no other sample amplitudes nearby, a small component along the direction of the second state would dominate the expansion coefficients at these points.  But, given those sample amplitudes, that would be the only way to represent such a component.
%\endinsert

@@

Let the singular value decomposition of $|A(α)〉$ be $|A(α)〉=|U〉SV†$.  The standard ways to compute singular values use the matrix
$$M(x,y)=\pmatrix{0 & |A'(x,y)〉\cr 〈A'(x,y)| & 0}.$$
Here, $|A(x,y)〉$ maps weight vectors in $C^R$ to kets, while $〈A(x,y)|=VS〈U|$ maps kets back to column vectors of inner products.  The operation of the zeros is obvious; formally, they need to be a row of null kets and a column of null bras.  The matrix $M(x,y)$ operates on a vector comprising a ket above $R$ complex numbers.  Now, the vector
$$x={1\over \sqrt2}\pmatrix{|u_R〉 \cr v_R}$$
is a normalised eigenvector  of $M$, belonging to the eigenvalue $σ_R(x,y)$.  As is well known in first-order pertubation theory, the derivative of this eigenvalue with respect to $x$ and $y$ is given by
$$∂_rσ_R=x†(∂_rM)x
	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA'〉v_R\right).$$
Substituting the partial derivatives gives
$$∂_rσ_R=\cases{\Re\left(〈u_R|(-x_r+a†)|α_r〉v_R^r\right) &  $1≤ r≤ R$, \cr
	\Re\left(〈u_R|(-y_r+ia†)|α_r〉v_R^r\right) & $R<r≤ 2R$.}$$

@@

%\topinsert 
%\vskip -40mm
%\XeTeXpicfile psfa.pdf width \hsize
%\vskip -90mm
%\XeTeXpicfile psv.pdf width \hsize\endinsert

@@

%\centerline{\XeTeXpicfile j1.pdf width \hsize}

%\centerline{\XeTeXpicfile j2.pdf width \hsize}

@@

%\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert

@@

%\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert

@@

%\topinsert \XeTeXpicfile spsns.pdf width \hsize
%Amplitudes of components \endinsert

%\topinsert \XeTeXpicfile traj.pdf width \hsize
%Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert

%\topinsert \XeTeXpicfile picard.pdf width \hsize
%Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert

%\topinsert \XeTeXpicfile svnum.pdf width \hsize
%Expected particle number for left singular vectors.  The svs with small sws have large particle number.  When $H$ is quartic, these have large components in $H|ψ〉$.  In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$. \endinsert

@@

%\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert

%\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert

@@

%\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}

%\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
